Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-19 main.Rmd
update before application

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. 
Now, some situations can lead to some of these sets being empty. We will discuss few cases, where one or more of these $nm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

- Data: <$C_1$, $C_2$, $v$>
- Aesthetic mapping (mapping of variables to elements of the plot): $C_1$ mapped to $x$ position and $v$ to $y$ position
- Statistical transformation (summarize data): Any descriptive or smoothing statistics that summarizes distribution of $v$
- Geometric objects (physical representation of the data): Any geometry displaying distribution, for example, boxplot, letter value, violin, ridge or highest density region plots
- Facet (split plots):  $C_2$

@@

If there are levels of cyclic granularities plotted across x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain emptry combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes.

@@

<!--In (b) we can not compare the distributions across different facets becuase every day-of-year cannot occur for every month. For example, 21st day of the year can never correspond to the month March or December of any year.-->
<!---In \autoref{fig:allFig} (c), there will be no observations for some combinations day-of-month and day-of-year. In particular, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. On the contrary, for \autoref{fig:allFig}--->  <!---Thus the graphs that don't work are those where many of the combination sets are empty.--->

@@

```{r allFig, fig.height=5.5, out.width="50%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Various probability distribution plots of electricity consumption data of Victoria from 2012 to 2014. (a) Letter value plot by day-of-month and month-of-year (b) Decile plot by day-of-year and month-of-year. Plot (a) and (b) show harmony and clashes respectively.", fig.show = 'hold'}

@@

VIC %>%
  create_gran("month_year") %>%
  filter(month_year %in% c("Jan", "Mar", "Dec")) %>% 
  create_gran("day_week") %>%
  filter(day_week %in% c("Sat", "Sun", "Mon")) %>%
  prob_plot("month_year", "day_week",
    response = "Demand",
    plot_type = "lv"
  ) +
  ylab("Electricity Demand [KWh]") +
  ggtitle("(a) Letter value plot with month-of-year facet")

@@

VIC %>%
  create_gran("month_year") %>%
  filter(month_year %in% c("Jan", "Mar", "Dec")) %>% 
  create_gran("day_week") %>%
  filter(day_week %in% c("Sat", "Sun", "Mon")) %>%
  prob_plot("day_week",
    "month_year",
    response = "Demand",
    plot_type = "lv"
  ) +
  ylab("") +
  xlab("Day of the Year") +
  ggtitle("(b) Letter value plot with day-of-week facet")

@@

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one `customer_id` shows a particular behavior, but this behavior is reversed when all keys are combined together. This is also intuitive because heterogenous `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. A strategy for analyzing multiple keys together could be to first group them basis time series or demographic features and then look at their energy behavior. This is beyond the scope of the current work.

>>>>

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets for each combination of levels ($A_i$, $B_j$) like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. 
Now, some situations can lead to few or many of these sets being empty. We will discuss few cases, where one or more of these $nm$ sets will be empty either due to the structure of the calendar, duration and location of events in a calendar or just by the construction of the cyclic granularities. <!--The implication of these empty sets on visualization will be discussed in Section \ref{sec:vis-interaction}. -->

@@

- Data: <$C_1$, $C_2$, $v$>
- Aesthetic mapping (mapping of variables to elements of the plot): $C_1$ mapped to $x$ position and $v$ to $y$ position
- Statistical transformation (data summarization): Any descriptive or smoothing statistics that summarizes distribution of $v$
- Geometric objects (physical representation of the data): Any geometry displaying distribution, for example, boxplot, letter value, violin, ridge or highest density region plots
- Facet (split plots):  $C_2$

@@

If there are levels of cyclic granularities plotted across x-axis which are not spanned by levels of cyclic granularities plotted across facets or vice versa, we will have empty sets leading to potential ineffective graphs. We hypothesize that the synergy of these cyclic granularities are thus playing a role while deciding if the resulting plot would be a good candidate for exploratory analysis. Harmonies are pairs of granularities that do not contain empty combinations and thus aid exploratory data analysis, and clashes are pairs that do not aid the process of exploratory analysis because they contain empty combinations. As a result, we should avoid plotting clashes.

@@

<!--In (b) we can not compare the distributions across different facets because every day-of-year cannot occur for every month. For example, 21st day of the year can never correspond to the month March or December of any year.-->
<!---In \autoref{fig:allFig} (c), there will be no observations for some combinations day-of-month and day-of-year. In particular, the 1st day of the month can never correspond to 2nd, 3rd or 4th day of the year. On the contrary, for \autoref{fig:allFig}--->  <!---Thus the graphs that don't work are those where many of the combination sets are empty.--->

@@

```{r allFig, fig.height=5.5, out.width="50%", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap="Letter value plot of electricity consumption data of Victoria from 2012 to 2014 (a) across day-of-week faceted by month-of-year and (b) across month-of-year faceted by day-of-week. Plots (a) and (b) both show harmonies since each month consists of all days of the week and every days of the week can occur in every month. However, Plot (a) helps comparison of different days of the week within each month and Plot (b) facilitates comparing months of the year for different days of the week.", fig.show = 'hold'}

@@

VIC %>%
  create_gran("month_year") %>%
  filter(month_year %in% c("Jan", "Mar", "Dec")) %>% 
  create_gran("day_week") %>%
  filter(day_week %in% c("Sat", "Sun", "Mon")) %>%
  prob_plot("month_year", "day_week",
    response = "Demand",
    plot_type = "lv"
  ) +
  ylab("Electricity Demand [KWh]") +
  xlab("day-of-week") + 
  ggtitle("(a) Letter value plot with month-of-year facet")

@@

VIC %>%
  create_gran("month_year") %>%
  filter(month_year %in% c("Jan", "Mar", "Dec")) %>% 
  create_gran("day_week") %>%
  filter(day_week %in% c("Sat", "Sun", "Mon")) %>%
  prob_plot("day_week",
    "month_year",
    response = "Demand",
    plot_type = "lv"
  ) +
  ylab("") +
  xlab("month-of-year") +
  ggtitle("(b) Letter value plot with day-of-week facet")

@@

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one `customer_id` shows a particular behavior, but this behavior is reversed when all keys are combined together. This is also intuitive because heterogeneous `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. A strategy for analyzing multiple keys together could be to first group them basis time series or demographic features and then look at their energy behavior. This is beyond the scope of the current work.
